---
title: "CPEB4_IP_Analysis"
output:
  rmdformats::readthedown:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    df_print: paged
    code_folding: show
---

> Analysis of CPEB4 binding to mRNAs in *in-vitro* NSCs (by RNA-immunoprecipitation = RIP) (Preprint: Goepferich & George et al. 2020)

Final plots in the paper may differ slightly due to:

* figures are not in order as in the paper (due to the intrinsic logic of this script and the input data)
* changes pending on the peer-review process
* collecting and final editing of sub-plots in Adobe Illustrator
* differences R package versions or other parameters

> Note: This is a reporting file providing all the necessary programming code to comprehend the main results of the paper.

> Note: Precompield files are provided in this directory.

```{r package-options, include=FALSE}
knitr::opts_knit$set(warning=FALSE, fig.width=8, fig.height=8)  
```

> Note: Computed as HTML-File

## Load the data (This will be provided in a table)

These are the RIP assay results (output from DESeq2)

```{r load_data}
load(file = 'rip_tbl.rda')

rip_tbl[ ,1:3]
```

> logFC_Enrichment = log-enrichment in CPEB4-IP fraction over control fractions (IGg and Input), n = 2

```{r packages, echo = FALSE, message = FALSE, warning = FALSE}
# load packages and set S4 class
require(tidyverse)
require(GenomicRanges)
require(ggpointdensity)
require(gridExtra)
require(ggrepel)
require(ggbeeswarm)
```

## Inspect the CPEB4-RIP assay as an MA-plot

```{r proteomics_volcano, warning=FALSE, fig.height=4, fig.width=6}

gg_cpeb4_rip <- ggplot(rip_tbl, aes(y = logFC_Enrichment, x =  Mean_IGg_Fraction)) +
  geom_point( data = filter(rip_tbl, is.na(Sig_Level_DESeq2) ), size = 1.5, color = 'darkgrey', alpha = .25) +
  geom_point( data = filter(rip_tbl, !is.na(Sig_Level_DESeq2) ),
              mapping = aes(color = Sig_Level_DESeq2), size = 1.5, alpha = .5) +
  geom_point( data = filter(rip_tbl, !is.na(Sig_Level_DESeq2) ), shape = 1, size = 1.5, alpha = .5) +
  geom_hline(yintercept = 0, col = 'black', size = .2, alpha = .25) +
  theme_minimal() +
  scale_x_log10( breaks = c(100, 1000, 10000), limits = c(25, 1e4) ) +
  ylim(c(-3,3)) +
  scale_colour_manual(values = c( 'firebrick4', 'red'), na.value = 'lightgrey') +
  labs(y = 'log enrichment ratio\n(Affinity of mRNA to CPEB4)',
       x = 'Read Counts in Control Samples',
       color = "Significance\n(RIP enriched)"
       ) +
  ggtitle('CPEB4 RIP Assay')

gg_cpeb4_rip

```


## End

```{r info}
sessionInfo()
```



```{r, include = FALSE, eval = FALSE}
# up-load to github
save(rip_tbl, file = 'rip_tbl.rda')
```





